\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-02 paper/ms.tex
Added section on multiplication and filtering

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,X^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $X$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,X^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,X^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,X^\T)\,D
    + D\,\mathrm{triu}(X\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,X^\T)\,D\,\mathrm{triu}(X\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $X$. \dfmtodo{X was used to represent independent
parameter matrix in definition of likelihood. Perhaps we should use W?}
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,X_{n-1,j}\,X_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
X_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
then these terms only add one to the rank of $U$ and $V$.  We use this fact
to simplify the algebra when real components are included:  if $J_{real}$
and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
the $U$, $V$, and $X$ matrices have size $N \times (J_{real}+2J_{complex})$.

@@

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
real number.
This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{X}_{n-1,j}\,z_{n-1}\right] \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{X}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

@@

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
\begin{equation}
{\bf y} = L D^{1/2} {\bf q}.
\end{equation}
This may be carried out in $O(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
\agoltodo{Check that these equations are in fact correct.}
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets; even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition 
and multiplication are used.

Multiplication of the full covariance matrix times a vector can be accomplished
by yet another multiplication by the transpose of the Cholesky decomposition.
\dfmtodo{Should we spell this out as well?  This is useful, for example, for
filter out a particular component of variability.}

@@

%Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.  With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $O(N+M)$ operations.

@@

Since the covariance matrix may be written as the sum of exponentials, we can write:
\begin{eqnarray}
\bar f_{*,i} &=& K(t_{*,i},{\bf t}) {\bf b} \\
&=& \sum_n \sum_p \alpha_p e^{-\beta_p \vert t_{*,i}-t_n\vert} b_j.
\end{eqnarray}
\agoltodo{Convert this to real, \celerite\ notation .}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
\begin{equation}
\bar f_{*,i} = \sum_{n=1}^{n_0} \sum_p \alpha_p e^{-\beta_p( t_{*,i}-t_n)} b_n + \sum_{n=n_0+1}^N \sum_p \alpha_p e^{-\beta_p(t_n-t_{*,i})} b_n.
\end{equation}
Each of these double-sums may have the sums interchanged, giving:
\begin{eqnarray}
\bar f_{*,i} &=& \sum_p e^{-\beta_p ( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} \alpha_p e^{-\beta_p ( t_{*,i-1}-t_n) } b_n \\
&+&
\sum_p e^{-\beta_p ( t_{*,i+1}-t_{*,i})} \sum_{n=n+0+1}^{N} \alpha_p e^{-\beta_p ( t_n-t_{*,i+1}) } b_n
\end{eqnarray}
We can compute this recursively.  Defining $Q^-_{i,n_0,p} = \sum_{j=1}^{n_0} \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ and defining $Q^+_{i,n_0,p} = \sum_{j=n_0}^N \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ gives:
\begin{equation}
\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
\end{equation}
Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $O(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx O(JM)$ operations
\begin{eqnarray}
Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
\end{eqnarray} 
for a total scaling of $\approx O(J^2N+J(N+M))$.

The variance evaluated at future predicted times ${\bf t_*} > {\bf t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},{\bf t}^T)K({\bf t},{\bf t})^{-1} K({\bf t},t_{*,i+1})\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
&\times& e^{-\beta_p(t_{*,i}-{\bf t}^T)}K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
\end{eqnarray}
for $1 \le p,q \le J$,
where
\begin{equation}
P_{i,p,q} = e^{-\beta_p(t_{*,i}-{\bf t}^T)}
K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}.
\end{equation}
For $i=1$, $P_{1,p,q}$ can be obtained in $O(J^2N)$ operations
with the Cholesky decomposition.
Subsequent $P_{i+1,p,q}$ may be obtained recursively as
\begin{equation}
P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
\end{equation}
in $O(J^2M)$ operations, for a total of $\approx O(J^2(N+M))$ operations.
Note that this variance formula {\it only} works for prediction; we have
yet to derive a formula for which ${\bf t_*}$ and ${\bf t}$ are
interspersed.

>>>>

To compute the marginalized likelihood defined in \eq{gp-likelihood}, we must
compute the log-determinant of $K$ and solve linear systems of the form
$K\,\bvec{b}=\bvec{y}$ for $\bvec{b}$, where $K$ is a rank-$2\,J$
semi-separable matrix.
If $K$ is positive semi-definite, these can be computed using the Cholesky
factorization of $K$:
\begin{eqnarray}
K = L\,D\,L^\T
\end{eqnarray}
where $L$ is a lower triangular matrix with unit diagonal and $D$ is a
diagonal matrix.
We use the ansatz that $L$ has the form
\begin{eqnarray}
    L = I + \mathrm{tril} (U\,W^T)
\end{eqnarray}
where $I$ is the identity, $U$ is the matrix from above, and $W$ is an unknown
$N \times 2\,J$ matrix.
Combining this with \eq{semi-sep}, we find the equation
\begin{eqnarray}\eqlabel{cholesky}
A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
&=& \left[I + \mathrm{tril} (U\,W^T)\right]\,D\,
    {\left[I + \mathrm{tril} (U\,W^T)\right]}^\T \\
&=& D + \mathrm{tril}(U\,W^\T)\,D
    + D\,\mathrm{triu}(W\,U^\T) \nonumber\\
&& +\, \mathrm{tril}(U\,W^\T)\,D\,\mathrm{triu}(W\,U^\T) \nonumber
\end{eqnarray}
that we can solve for $W$.
Setting each element on the left side of \eq{cholesky} equal to the
corresponding element on the right side, we derive the following recursion
relations
\begin{eqnarray}
S_{n,j,k} &=& S_{n-1,j,k} + D_{n-1,n-1}\,W_{n-1,j}\,W_{n-1,k} \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} U_{n,j}\,S_{n,j,k}\,U_{n,k}
    \nonumber\\
W_{n,j} &=& \frac{1}{D_{n,n}}\left[ V_{n,j} -
    \sum_{k=1}^{2\,J}U_{n,k}\,S_{n,j,k} \right]
    \eqlabel{algorithm}
\end{eqnarray}
where $S_n$ is a symmetric $2\,J \times 2\,J$ matrix and every element
$S_{1,j,k}=0$.
The computational cost of one iteration of the update in \eq{algorithm} is
$\mathcal{O}(J^2)$ and this must be run $N$ times~--~once for each
observation~--~so the total cost of this factorization scales as
$\mathcal{O}(N\,J^2)$.

Note that if some \celeriteterm\ terms are real, with $b_j=d_j=0$,
then these terms only add one to the rank of $U$ and $V$.  We use this fact
to simplify the algebra when real components are included:  if $J_{real}$
and $J_{complex}$ terms are included, with $J_{real}+J_{complex}=J$, then
the $U$, $V$, and $W$ matrices have size $N \times (J_{real}+2J_{complex})$.

@@

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
real number.
This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{W}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,W_{n,2\,j-1} \\
\tilde{W}_{n,2\,j} &=& e^{-c_j\,t_n}\,W_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{W}_{n-1,j}\,\tilde{W}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{W}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

The inverse of $K$ can be applied using the relationship
\begin{eqnarray}
\bvec{b} = K^{-1}\,\bvec{y} = {(L^\T)}^{-1}\,D^{-1}\,L^{-1}\,\bvec{y} \quad.
\end{eqnarray}
First, we solve for $\bvec{z} = L^{-1}\,\bvec{y}$ using the following forward
substitution algorithm
\begin{eqnarray}
    f_{n,j} &=& \phi_{n,j}\,\left[f_{n-1,j} +
    \tilde{W}_{n-1,j}\,z_{n-1}\right] \nonumber\\
    z_n &=& y_n - \sum_{j=1}^{2\,J} \tilde{U}_{n,j}\,f_{n,j}
\end{eqnarray}
where $f_{0,j} = 0$ for all $j$.
Using this result and the following backward substitution algorithm, we can
compute $\bvec{b}$
\begin{eqnarray}
g_{n,j} &=& \phi_{n+1,j}\,\left[g_{n+1,j} +
    \tilde{U}_{n+1,j}\,b_{n+1}\right] \nonumber\\
b_n &=& \frac{z_n}{D_{n,n}} - \sum_{j=1}^{2\,J} \tilde{W}_{n,j}\,g_{n,j}
\end{eqnarray}
where $g_{N+1,j} = 0$ for all $j$.
The total cost of the forward and backward substitution passes scales as
$\mathcal{O}(N\,J)$.

@@

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
\begin{equation}
{\bf y} = L D^{1/2} {\bf q}.
\end{equation}
This may be carried out in $\mathcal{O}(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{W}_{n-1,j} D_{n-1,n-1}^{1/2} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
\agoltodo{Check that these equations are in fact correct.}
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets; even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition 
and multiplication are used.

\subsection{Multiplication and filtering}

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(JN)$
operations with a stable recursive algorithm using the semi-separable form:
\begin{equation}
{\bf y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) {\bf z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \sum_{j=1}^J \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \sum_{j=1}^J \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^J \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,J$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards 
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.  \dfmtodo{I'm getting confused by zero-based
indexing in your C code...  this may be worth scrutinizing to see if I screwed up.}

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K({\bf t},{\bf t}) \left[K({\bf t},{\bf t})+\sigma^2{\bf I}\right]^{-1} {\bf y},\\
&=& K({\bf t},{\bf t}) {\bf b}
\end{eqnarray}
where ${\bf y}$ is the vector of measurements and ${\bf b}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times ${\bf b}$ may be accomplished in $\mathcal{O}(JN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = {\bf y} -\sigma^2 {\bf I} {\bf b}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.

@@

%Gaussian process regression may also be used for prediction and interpolation at future or missing data, based upon current measurements.  With \celerite\ it proves possible to predict the mean value at times ${t_m; m=1,...,M}$ based upon measured times ${t_n; n=1,...,N}$ in $\mathcal{O}(N+M)$ operations.

@@

Since the covariance matrix may be written as the sum of exponentials, we can write:
\begin{eqnarray}
\bar f_{*,i} &=& K(t_{*,i},{\bf t}) {\bf b} \\
&=& \sum_n \sum_p \alpha_p e^{-\beta_p \vert t_{*,i}-t_n\vert} b_j.
\end{eqnarray}
\agoltodo{Convert this to real, \celerite\ notation .}
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,i}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,i}$, gives
\begin{equation}
\bar f_{*,i} = \sum_{n=1}^{n_0} \sum_p \alpha_p e^{-\beta_p( t_{*,i}-t_n)} b_n + \sum_{n=n_0+1}^N \sum_p \alpha_p e^{-\beta_p(t_n-t_{*,i})} b_n.
\end{equation}
Each of these double-sums may have the sums interchanged, giving:
\begin{eqnarray}
\bar f_{*,i} &=& \sum_p e^{-\beta_p ( t_{*,i}-t_{*,i-1})} \sum_{n=1}^{n_0} \alpha_p e^{-\beta_p ( t_{*,i-1}-t_n) } b_n \\
&+&
\sum_p e^{-\beta_p ( t_{*,i+1}-t_{*,i})} \sum_{n=n+0+1}^{N} \alpha_p e^{-\beta_p ( t_n-t_{*,i+1}) } b_n
\end{eqnarray}
We can compute this recursively.  Defining $Q^-_{i,n_0,p} = \sum_{j=1}^{n_0} \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ and defining $Q^+_{i,n_0,p} = \sum_{j=n_0}^N \alpha_p e^{-\beta_p (t_{*,i}-t_j)} b_j$ gives:
\begin{equation}
\bar f_{*,i} = \sum_p e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i-1,n_0,p} + \sum_p e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i+1,n_0,p}.
\end{equation}
Once $Q^-_{1,p}$ and $Q^+_{N,p}$ are obtained in $\mathcal{O}(JN)$ operations, then the subsequent $Q^-_{i,p}$ and $Q^+_{i,p}$ values can be obtained in $\approx \mathcal{O}(JM)$ operations
\begin{eqnarray}
Q^-_{i,p} &=& e^{-\beta_p (t_{*,i}-t_{*,i-1})} Q^-_{i,p},\\
Q^+_{i,p} &=& e^{-\beta_p (t_{*,i+1}-t_{*,i})} Q^+_{i,p},
\end{eqnarray} 
for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.

The variance evaluated at future predicted times ${\bf t_*} > {\bf t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},{\bf t}^T)K({\bf t},{\bf t})^{-1} K({\bf t},t_{*,i+1})\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
&\times& e^{-\beta_p(t_{*,i}-{\bf t}^T)}K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}\\
&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
\end{eqnarray}
for $1 \le p,q \le J$,
where
\begin{equation}
P_{i,p,q} = e^{-\beta_p(t_{*,i}-{\bf t}^T)}
K({\bf t},{\bf t})^{-1} e^{-\beta_q(t_{*,i}-{\bf t})}.
\end{equation}
For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
with the Cholesky decomposition.
Subsequent $P_{i+1,p,q}$ may be obtained recursively as
\begin{equation}
P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
\end{equation}
in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
Note that this variance formula {\it only} works for prediction; we have
yet to derive a formula for which ${\bf t_*}$ and ${\bf t}$ are
interspersed.
